Quantile Trend Filtering via Spingarn's Method

### 2017-06-02 Notes/detrendify_spingarn.tex
built pdf

The classic least squares regression is notoriously sensitive to outliers. One remedy to blunt the influence of outliers is to compute the least absolute deviations (LAD) solution in place of the least squares one. Given a design matrix $\M{X} \in \Real^{n \times p}$ and continuous responses $\V{y} \in \Real^n$, we estimate a regression vector $\V{\theta} \in \Real^p$ so that $\M{X}\V{\theta}$ is a good approximation of $\V{y}$. The LAD estimator is a solution to the problem
\begin{eqnarray*}
\underset{\V{\theta}}{\min}\; \frac{1}{n}\lVert \V{y} - \M{X}\V{\theta} \rVert_1.
\end{eqnarray*}
The above optimization problem generalizes the notion of the median of a collection of numbers. A median $\mu$ of $n$ reals $\VE{y}{1}, \ldots, \VE{y}{n}$ is the minimizer of the function
\begin{eqnarray*}
f(u) & = & \frac{1}{n}\sum_{i=1}^n \lvert \VE{y}{i} - \theta \rvert.
\end{eqnarray*}
Recall that the median is the 50th percentile or 0.5-quantile, namely half of the $\VE{y}{i}$ are less than or equal to $\mu$ and the other half is greater than or equal to $\mu$. The median can be generalized to arbitrary $\tau$-quantiles for $\tau \in (0,1)$ as follows. First define the so-called ``check function"
\begin{eqnarray*}
\rho_\tau(\Delta) & = & \begin{cases}
\tau \Delta & \text{$\Delta \geq 0$} \\
-(1-\tau)\Delta & \text{$\Delta < 0$}. \\
\end{cases}
\end{eqnarray*}
Then the $\tau$th quantile of the $\VE{y}{i}$ is a minimizer of the function
\begin{eqnarray*}
f_\tau(\theta) & = & \frac{1}{n}\sum_{i=1}^n \rho_\tau(\VE{y}{i} - \theta).
\end{eqnarray*}
Returning to the regression context, we can generalize LAD regression to quantile regression, namely computing the minimizer of the function
\begin{eqnarray*}
f_\tau(\V{\theta}) & = & \frac{1}{n}\sum_{i=1}^n \rho_\tau(\VE{y}{i} - \langle \V{x}_i \mid \V{\theta} \rangle),
\end{eqnarray*}
where $\V{x}_i \in \Real^p$ denotes the $i$th row of $\M{X}$.

@@

\begin{itemize}
	\item Read compressed data
	\item May want to read data directly in C and not pull into R; make R just an interface
	\item Use historical data to choose $\tau$ and $\lambda$
	\item 
\end{itemize}

>>>>

The classic least squares regression is notoriously sensitive to outliers. One remedy to blunt the influence of outliers is to compute the least absolute deviations (LAD) solution in place of the least squares one. Given a design matrix $\M{X} \in \Real^{n \times p}$ and continuous responses $\V{y} \in \Real^n$, we estimate a regression vector $\V{\theta} \in \Real^p$ so that $\M{X}\V{\theta}$ is a good approximation of $\V{y}$. The LAD estimator is a solution to the problem
\begin{eqnarray*}
\underset{\V{\theta}}{\min}\; \frac{1}{n}\lVert \V{y} - \M{X}\V{\theta} \rVert_1.
\end{eqnarray*}
The above optimization problem generalizes the notion of the median of a collection of numbers. A median $\mu$ of $n$ reals $\VE{y}{1}, \ldots, \VE{y}{n}$ is the minimizer of the function
\begin{eqnarray*}
f(u) & = & \frac{1}{n}\sum_{i=1}^n \lvert \VE{y}{i} - \theta \rvert.
\end{eqnarray*}
Recall that the median is the 50th percentile or 0.5-quantile, namely half of the $\VE{y}{i}$ are less than or equal to $\mu$ and the other half is greater than or equal to $\mu$. The median can be generalized to arbitrary $\tau$-quantiles for $\tau \in (0,1)$ to give us quantile regression \cite{Koenker1978}.

First define the so-called ``check function"
\begin{eqnarray*}
\rho_\tau(\Delta) & = & \begin{cases}
\tau \Delta & \text{$\Delta \geq 0$} \\
-(1-\tau)\Delta & \text{$\Delta < 0$}. \\
\end{cases}
\end{eqnarray*}
Then the $\tau$th quantile of the $\VE{y}{i}$ is a minimizer of the function
\begin{eqnarray*}
f_\tau(\theta) & = & \frac{1}{n}\sum_{i=1}^n \rho_\tau(\VE{y}{i} - \theta).
\end{eqnarray*}
Returning to the regression context, we can generalize LAD regression to quantile regression, namely computing the minimizer of the function
\begin{eqnarray*}
f_\tau(\V{\theta}) & = & \frac{1}{n}\sum_{i=1}^n \rho_\tau(\VE{y}{i} - \langle \V{x}_i \mid \V{\theta} \rangle),
\end{eqnarray*}
where $\V{x}_i \in \Real^p$ denotes the $i$th row of $\M{X}$.

@@

\begin{itemize}
	\item Read compressed data
	\item May want to read data directly in C and not pull into R; make R just an interface
	\item Use historical data to choose $\tau$ and $\lambda$
\end{itemize}

\section{To Do}

\begin{itemize}
	\item Find out if we can use the EPA data. Add co-authors?
	\item Make version 2 with the three fast fused lasso solvers
	\item Add homotopy / warm start
	\item Add model selection
	\item Do experiments to evaluate different choices of $t$ and $\lambda^{(k)}$
	\item See if Anderson acceleration helps. Talk to Tim Kelley?
	\item Write vignette
	\item Do comparisons with BEAD
	\item Do wind polar plots with and without removing trends
	\item Compare with Splines
\end{itemize}
